---
title: "Breitenbach 2.0 - DIPTERA"
author: "Viktor Baranov & Nathan Jay Baker"
editor: visual
date: 2024/12/18
format:
  html: 
    code-overflow: scroll
    fig-height: 12
    fig-width: 12
    fig-format: svg
    # pdf-engine: lualatex
    eval: TRUE
    warning: false
---

# Load necessary packages

```{r}
library(tidyverse)
library(patchwork)
library(janitor)
library(SpadeR)
library(vegan)
library(codyn)
library(mobr)
library(iNEXT)
library(readxl)
library(writexl)
```

# Load the Dipteran data

```{r}
diptera <- read_xlsx("Data/Breitenbach_community_data_17.12.2024.xlsx") |> # read in excel spreadsheet
  as_tibble() |>                                                           # Make tibble table
  clean_names() |>                                                         # Homogenize column names
  rename_with(~ gsub("^x", "", .x), .cols = matches("^x[0-9]")) |>         # Change column names
  filter(!str_detect(original_name, "Summe")) |>                           # delete rows containing "summe"
  filter(order == "Diptera") |>                                            # Keeps only Dipteran taxa
  select(where(~ !is.numeric(.x) || sum(.x) != 0)) |>                      # Removes empty (0) columns
  mutate(trap_new = case_when(                                             # Homogenize trap names
    trap %in% c('Haus 0') ~ 'O',
    trap %in% c('Haus A', 'A/I') ~ 'A',
    trap %in% c('Haus I') ~ 'I',
    trap %in% c('Haus B', 'B / II', 'B-II-X', 'B/II', 'Haus B/II') ~ 'B',
    trap %in% c('Haus C-IV', 'C / IV', 'C-IV', 'C/IV', 'Haus C', 'Haus C/IV') ~ 'C',
    trap %in% c('Haus III') ~ 'III',
    trap %in% c('Haus D') ~ 'D',
    trap %in% c('Haus E - V', 'E / V', 'E-V', 'E/V', 'Haus E/V') ~ 'E',
    trap %in% c('Haus F - VI', 'F / VI', 'F-VI', 'F-VII', 'F/VI', 'Haus F/VI') ~ 'F',
    trap %in% c('Haus G', 'G-VII', 'G/ VII', 'G/VII', 'Haus G/VII') ~ 'G',
    trap %in% c('Quelle') ~ 'Source',
    TRUE ~ trap)) |>
  select(-c(taxa_id:trap)) |>                                              # Remove unnecessary columns
  mutate(sp_name  = validated_name,                                        # Change column names
         trap     = trap_new,                                              # Change column names
         trap     = factor(trap)) |>                                       # make trap a factor
  arrange(trap, family, sp_name) |>                                        # Sort data
  select(-c(validated_name, trap_new, order)) |>                           # Remove unnecessary columns
  select(trap, sp_name, family, everything()) |>                           # rearrange columns
  print()
```

# Initial data wrangling

```{r}
diptera_clean <- diptera |>
  group_by(trap,
           sp_name,
           family) |>                                                      # Grouping by trap, sp_name, and family
  summarise(across(where(is.numeric),
                   ~sum(.x, na.rm = TRUE)),
                   .groups = "drop")

diptera_long <- diptera_clean |>
  pivot_longer(
    cols = starts_with("19") | starts_with("20"),                          # Specify the year column to pivot
    names_to = "year",                                                     # Name column that will hold year values
    values_to = "abundance") |>                                            # Name column that will hold the counts
  mutate(trap_code = paste(trap, year, sep = "_")) |>                      # create new column with trap_code (trap name + year)
  mutate(year = as.numeric(year)) |>                                       # make year variable numeric
  filter(abundance != 0) |>                                                # Remove rows where count is 0
  arrange(trap, year)                                                      # Sort data

diptera_agg <- diptera_long |>
  group_by(sp_name,
           trap) |>                                                        # Group by sp_name and trap
  summarise(abundance = sum(abundance),
            .groups = "drop") |>                                           # Summarise with the sum of "abundance"
  arrange(trap)                                                            # Sort data

diptera_wide <- diptera_agg |>
  pivot_wider(names_from  = trap,
              values_from = abundance,
              values_fill = 0,
              values_fn   = sum) |>
  arrange(sp_name)                                                         # Sort data
```

# Calculating indices

## Chao's indices: species estimation in a single community

```{r}
chaoA_dipt      <- ChaoSpecies(diptera_wide$A ,"abundance", k = 2, conf = 0.95)      # Diptera trap A
chaoB_dipt      <- ChaoSpecies(diptera_wide$B ,"abundance", k = 2, conf = 0.95)      # Diptera trap B
chaoC_dipt      <- ChaoSpecies(diptera_wide$C ,"abundance", k = 2, conf = 0.95)      # Diptera trap C
chaoD_dipt      <- ChaoSpecies(diptera_wide$D ,"abundance", k = 2, conf = 0.95)      # Diptera trap D
chaoE_dipt      <- ChaoSpecies(diptera_wide$E ,"abundance", k = 2, conf = 0.95)      # Diptera trap E
chaoF_dipt      <- ChaoSpecies(diptera_wide$F ,"abundance", k = 2, conf = 0.95)      # Diptera trap F
chaoG_dipt      <- ChaoSpecies(diptera_wide$G ,"abundance", k = 2, conf = 0.95)      # Diptera trap G
chaoI_dipt      <- ChaoSpecies(diptera_wide$I ,"abundance", k = 2, conf = 0.95)      # Diptera trap I
chaoIII_dipt    <- ChaoSpecies(diptera_wide$III ,"abundance", k = 2, conf = 0.95)    # Diptera trap III
```

## Standard community indices + turnover

```{r}
diptera_long$ro.ab <- round(diptera_long$abundance, digits = 0)                      # Round abundance values to whole numbers for rarefaction analysis
TD <- NULL                                                                           # Initialize empty data frame to store results

for (i in unique(diptera_long$trap)) {                                               # Loop through each unique trap in the dataset
  sub <- diptera_long[diptera_long$trap == i, ]                                      # Create subset for current trap and reshape data from long to wide format
  sub_m <- sub |>
    select(trap_code, sp_name, abundance) |>
    pivot_wider(names_from = sp_name, values_from = abundance, values_fill = 0)
  sub_ta <- sub_m[, -1]                                                              # Remove trap_code column for calculations

  # Calculate diversity indices
  SppRich <- specnumber(sub_ta)                                                      # Species richness (total number of species)
  Simp <- diversity(sub_ta, index = "simpson")                                       # Simpson's diversity (probability two random individuals are different species)
  Shan <- diversity(sub_ta, index = "shannon")                                       # Shannon's diversity (accounts for both abundance and evenness)
  EvenJ <- Shan / log(SppRich)                                                       # Pielou's evenness (how close in numbers each species is)
  E10 <- Shan / SppRich                                                              # Shannon's evenness (alternative evenness measure)
  Abund <- rowSums(sub_ta)                                                           # Total abundance (sum of all individuals)
  S_PIE <- calc_PIE(sub_ta, ENS = TRUE)                                              # Effective number of common species

  # Calculate species turnover metrics between years
  DATA1_Turnover <- codyn::turnover(sub,
    time.var = "year",
    species.var = "sp_name",
    abundance.var = "abundance",
    replicate.var = NA,
    metric = "total")                                                                # Total turnover (appearances + disappearances)
  Turnover <- c("NA", DATA1_Turnover$total)                                          # Add NA for first year

  DATA1_Turnover_app <- codyn::turnover(sub,                                         # Species appearances only
    time.var = "year",
    species.var = "sp_name",
    abundance.var = "abundance",
    replicate.var = NA,
    metric = "appearance")
  Turnover_app <- c("NA", DATA1_Turnover_app$appearance)

  DATA1_Turnover_disapp <- codyn::turnover(sub,                                      # Species disappearances only
    time.var = "year",
    species.var = "sp_name",
    abundance.var = "abundance",
    replicate.var = NA,
    metric = "disappearance")
  Turnover_disapp <- c("NA", DATA1_Turnover_disapp$disappearance)

  # Prepare data for rarefaction analysis
  sub_m_r <- sub |>
    select(trap_code, sp_name, ro.ab) |>
    pivot_wider(names_from = sp_name,
      values_from = ro.ab,
      values_fill = 0)                                                               # Create matrix with rounded abundances
  sub_ta_r <- sub_m_r[, -1]                                                          # Remove trap_code column

  # Calculate rarefied species richness
  rare.SppRich <- if (min(rowSums(sub_ta_r)) > 10) {
    rarefy(sub_ta_r, sample = min(rowSums(sub_ta_r)))
  } else {
    rarefy(sub_ta_r, sample = 10)
  }                                                                                  # If minimum sample size > 10, use that; otherwise use 10 as minimum

  # Combine all calculated metrics into a data frame
  TD.i <- data.frame(sub_m$trap_code,
    SppRich,
    Simp,
    Shan,
    EvenJ,
    E10,
    Abund,
    S_PIE,
    Turnover,
    Turnover_app,
    Turnover_disapp,
    rare.SppRich)

  # Append results to main data frame
  TD <- rbind(TD, TD.i)

  # Clean up temporary variables to avoid conflicts in next iteration
  rm(TD.i, sub_m, sub_ta, sub, SppRich, Simp, Shan, EvenJ, E10, Abund, S_PIE,
    DATA1_Turnover, Turnover, DATA1_Turnover_app, Turnover_app,
    DATA1_Turnover_disapp, Turnover_disapp, sub_m_r, sub_ta_r, rare.SppRich)
}

# Clean up dataframe
TD <- TD |>
  as_tibble() |>                                                                     # Convert to tibble format for better handling
  rename(trap_code = sub_m.trap_code) |>                                             # Fix column name from earlier processing
  mutate(trap = gsub("_.*", "", trap_code),                                          # Remove everything after _ to get trap ID
         year = gsub(".*_", "", trap_code)) |>                                       # Remove everything before _ to get year
  mutate(across(-c(trap_code, trap), as.numeric)) |>                                 # Convert columns to numeric format
  select(trap_code, trap, year, everything()) |>                                     # Reorder columns: trap_code, trap, year first, then rest
  print()
```

## Plotting

### Define plotting parameters

```{r}
My_theme <- theme(panel.background = element_blank(),
                  panel.border = element_rect(fill = NA, linewidth = 1.25),
                  strip.background = element_rect(fill = "white", 
                                                  color = "white", linewidth = 1.25),
                  legend.position = "bottom",
                  text = element_text(size = 16))
```

### Turnover (total) by trap by year

```{r}
ggplot(TD |> 
         group_by(trap) |> 
         filter(n() > 3),
       aes(x = year, y = Turnover)) +
  geom_line() +      
  geom_smooth(method = "lm") +  
  geom_smooth(method = "gam", col = "red") +
  geom_point(size = 2) +   
  geom_hline(yintercept = 0.5,            
             linetype = "dotted",           
             color = "grey50",               
             size = 0.5) +    
  facet_wrap(~trap) +                        
  scale_x_continuous(breaks = seq(min(TD$year), max(TD$year), by = 3)) +
  scale_y_continuous(limits = c(0, max(TD$Turnover, na.rm = TRUE))) +  
  labs(title = "Turnover",  
       x = "Year",                            
       y = "Turnover",
       caption = "Species turnover over time by trap") +                      
  My_theme +                     
  My_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.ticks.length = unit(0.25, "cm"))
```

### Turnover (appearances) by trap by year

```{r}
ggplot(TD |> 
         group_by(trap) |> 
         filter(n() > 3),
       aes(x = year, y = Turnover_app)) +
  geom_line() +      
  geom_smooth(method = "lm") +  
  geom_smooth(method = "gam", col = "red") +
  geom_point(size = 2) +   
  geom_hline(yintercept = 0.5,            
             linetype = "dotted",           
             color = "grey50",               
             size = 0.5) +    
  facet_wrap(~trap) +                        
  scale_x_continuous(breaks = seq(min(TD$year), max(TD$year), by = 3)) +
  scale_y_continuous(limits = c(0, max(TD$Turnover, na.rm = TRUE))) +  
  labs(title = "Turnover: appearences",  
       x = "Year",                            
       y = "Turnover appearences",
       caption = "Species appearences over time by trap") +                      
  My_theme +                     
  My_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.ticks.length = unit(0.25, "cm"))
```

### Turnover (disappearances) by trap by year

```{r}
ggplot(TD |> 
         group_by(trap) |> 
         filter(n() > 3),
       aes(x = year, y = Turnover_disapp)) +
  geom_line() +   
  geom_smooth(method = "lm") +  
  geom_smooth(method = "gam", col = "red") +
  geom_point(size = 2) +   
  geom_hline(yintercept = 0.5,            
             linetype = "dotted",           
             color = "grey50",               
             size = 0.5) +    
  facet_wrap(~trap) +                        
  scale_x_continuous(breaks = seq(min(TD$year), max(TD$year), by = 3),
                     minor_breaks = seq(min(TD$year), max(TD$year), by = 1)) +
  scale_y_continuous(limits = c(0, max(TD$Turnover, na.rm = TRUE))) +  
  labs(title = "Turnover: disappearences",  
       x = "Year",                            
       y = "Turnover Disappearences",
       caption = "Species dissappearences through time by traps") +                      
  My_theme +                    
  My_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.ticks.length = unit(0.25, "cm"))
```

### Abundance by trap by year

```{r}
ggplot(TD |> 
         group_by(trap) |> 
         filter(n() > 3),
       aes(x = year, y = Abund)) +
  geom_line() +   
  geom_smooth(method = "lm") +  
  geom_smooth(method = "gam", col = "red") +
  geom_point(size = 2) +   
  geom_hline(yintercept = 0.5,            
             linetype = "dotted",           
             color = "grey50",               
             size = 0.5) +    
  facet_wrap(~trap, scales = "free_y") +                        
  scale_x_continuous(breaks = seq(min(TD$year), max(TD$year), by = 3)) +
  labs(title = "Abundance",  
       x = "Year",                            
       y = "Abundance",
       caption = "Abundance through time by trap") +                      
  My_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.ticks.length = unit(0.25, "cm")) 
```

### Log abundance (log10 + 1) by trap by year

```{r}
ggplot(TD |> 
         group_by(trap) |> 
         filter(n() > 3),
       aes(x = year, y = log(Abund + 1))) +
  geom_line() +   
  geom_smooth(method = "lm") +  
  geom_smooth(method = "gam", col = "red") +
  geom_point(size = 2) +   
  # geom_hline(yintercept = 0.5,
  #            linetype = "dotted",
  #            color = "grey50",
  #            size = 0.5) +
  facet_wrap(~trap, scales = "free_y") +      
  scale_y_continuous(breaks = seq(min(0), max(10), by = 2)) +
  scale_x_continuous(breaks = seq(min(TD$year), max(TD$year), by = 3)) +
  labs(title = "Log abundance",  
       x = "Year",                            
       y = "log(abundance + 1)",
       caption = "Log abundance through time by trap") +                      
  My_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.ticks.length = unit(0.25, "cm")) 

```

### Species richness by trap by year

```{r}
ggplot(TD |> 
         group_by(trap) |> 
         filter(n() > 3),
       aes(x = year, y = SppRich)) +
  geom_line() +   
  geom_smooth(method = "lm") +  
  geom_smooth(method = "gam", col = "red") +
  geom_point(size = 2) +   
  # geom_hline(yintercept = 0.5,            
  #            linetype = "dotted",           
  #            color = "grey50",               
  #            size = 0.5) +    
  facet_wrap(~trap, scales = "free_y") +                        
  scale_x_continuous(breaks = seq(min(TD$year), max(TD$year), by = 3)) +
  labs(title = "Species richness",  
       x = "Year",                            
       y = "Species richness",
       caption = "Species richness through time by trap") +                      
  My_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.ticks.length = unit(0.25, "cm")) 
```

### (Individual-based) rarefied species richness by trap by year

```{r}
ggplot(TD |> 
         group_by(trap) |> 
         filter(n() > 3),
       aes(x = year, y = rare.SppRich)) +
  geom_line() +   
  geom_smooth(method = "lm") +  
  geom_smooth(method = "gam", col = "red") +
  geom_point(size = 2) +   
  # geom_hline(yintercept = 0.5,            
  #            linetype = "dotted",           
  #            color = "grey50",               
  #            size = 0.5) +    
  facet_wrap(~trap, scales = "free_y") +                        
  scale_x_continuous(breaks = seq(min(TD$year), max(TD$year), by = 3)) +
  labs(title = "Rarefied species richness",  
       x = "Year",                            
       y = "Rarefied species richness",
       caption = "Rarefied species richness through time by trap") +                      
  My_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.ticks.length = unit(0.25, "cm")) 
```

# Rarefaction accumulation curves

## Calculation of rarefaction curves

```{r}
# diptera_wide_red <- diptera_wide |>
#   select(-D, -F, -Source) |>
#   column_to_rownames(var = "sp_name") |>
#   data.frame() |>
#   print()                                                                             # remove traps with lower quality data: Trap D, F, & Quelle
# diptera_cum.rare = iNEXT(diptera_wide_red, q = 0, datatype = "abundance")             # calculate accumulation curves
# write_rds(diptera_cum.rare, "Outputs/rarefaction_diptera.rds")                        # save output for quicker loading next time
diptera_cum.rare <- readRDS("Outputs/rarefaction_diptera.rds")                          # load saved output
```

### Check outputs

#### Basic information

```{r}
diptera_cum.rare$DataInfo
```

#### Asymptotic diversity estimates.

```{r}
diptera_cum.rare$AsyEst
```

## Plotting

### Plotting parameters

```{r}
My_theme <- theme(panel.background = element_blank(),
                  panel.border = element_rect(fill = NA, linewidth = 1.25),
                  strip.background = element_rect(fill = "white", 
                                                  color = "white", linewidth = 1.25),
                  legend.position = "right",
                  text = element_text(size = 16),
                  axis.text.x = element_text(angle = 45, hjust = 1))
```

### All accumulation curves in one plot

```{r}
options(scipen = 999) # Increase penalty to avoid scientific notation
diptera_p <- ggiNEXT(diptera_cum.rare, type = 1, color.var = "Assemblage")
diptera_p <- diptera_p + ggtitle("Diptera")
diptera_p <- diptera_p + theme(legend.position = "bottom")
diptera_p
```

### Accumulation curves in separate plots

```{r}
options(scipen = 999) # Increase penalty to avoid scientific notation
diptera_p1 <- ggiNEXT(diptera_cum.rare, facet.var = "Assemblage", grey = T)
diptera_p1 <- diptera_p1 + ggtitle("Diptera")
diptera_p1 <- diptera_p1 + theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom")
diptera_p1
```
